---
title: "Project1_Group2_writeup"
author: "Michael Sarkis,,,"
date: "8/31/2021"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
```


# Analysis Plan

## Data

Income_Unemployment - Median County Income and Unemployment rates
Urbanization_Index - County level urbanization measure
Vaccinations - County level vaccination data
us-counties - Cases, Deaths and Hospitalizations

## Analysis Ideas


1. Predicting Covid Cases by Median-Income, Vaccination Rate, Urbanization

2. Predicting Mask Use by Median Income, Urbanization, Vaccination Rate

TASKS TO DO

0. Merge Data

1. Decide on final goal

2. Decide on data to use

3. Discuss possible models

##  Merge data

NA values for FIPS code in cases_death

```{r read_data}
cases_deaths = read.csv("data/us-counties.csv")[,1:5]
income_unemp = read_excel("data/Income_Unemployment.xlsx", skip = 4)[,c(1,4,88,89,90,91)]
unemp2021 = read.table("data/labor_force_latest.txt", 
                       skip = 6, sep = "|", fill = TRUE,
                       col.names = c("area code", "FIPS1", "FIPS2", "area title",
                                     "period", "civilian labor force", "employed",
                                     "unemployed", "unemployed rate",
                                     "median.income"))[,c(2,3,5,7,8,9)]
voting = read.csv("data/voting_data.csv")[,c(2,5,6,9)]
region = read.table("data/us_regions.txt", sep = ",", header = TRUE)[,c(1,3)]
geographic = read.csv("data/uscounties.csv")[,c(4,7,8,9)]
```

```{r fips cleaning}
#get same format and name for FIPS code, date and year
cases_deaths$fips = formatC(cases_deaths$fips, width = 5, format = "d", flag = "0")
cases_deaths$date = as.Date(cases_deaths$date)

names(income_unemp) = c("fips","urban_code","employed","unemployed",
                        "unemployed.rate","median.income")
income_unemp$year = rep("2020", nrow(income_unemp))
income = income_unemp[,c(1,2,6)]
unemp2020 = income_unemp[,c(1,3:5,7)]

unemp2021 = unemp2021[unemp2021$period == "   May-21  ",]
unemp2021$FIPS1 = formatC(unemp2021$FIPS1, width = 2, format = "d", flag = "0")
unemp2021$FIPS2 = formatC(unemp2021$FIPS2, width = 3, format = "d", flag = "0")
unemp2021$fips =paste(unemp2021$FIPS1, unemp2021$FIPS2, sep = "")
unemp2021 = unemp2021[,c(7,4:6)]
unemp2021$year = rep("2021", nrow(unemp2021))

voting$county_fips = formatC(voting$county_fips, width = 5, format = "d", flag = "0")
names(voting)[1] = "fips"

names(region)[1] = "state"

geographic$county_fips = formatC(geographic$county_fips, width = 5, format = "d", flag = "0")
names(geographic)[1] = "fips"
```


```{r combining nyc}
cases_deaths[cases_deaths$county == "New York City",]$fips = "nyc"

nyc = c("36005","36047","36061","36081","36085")
geographic_nyc = geographic %>% 
  filter(fips %in% nyc) %>%  
  summarize(lat = mean(lat), lng = mean(lng), population = sum(population))
geographic_nyc$fips = "nyc"
geographic_nyc = geographic_nyc[,c(4,1:3)]
geographic = geographic %>% 
  filter(!fips %in% nyc)
geographic = rbind(geographic, geographic_nyc)

income_nyc = income %>% 
  filter(fips %in% nyc) %>%  
  summarize(median.income = mean(median.income))
income_nyc$fips = "nyc"
income_nyc$urban_code= 1
income_nyc = income_nyc[,c(2,1,3)]
income = income %>% 
  filter(!fips %in% nyc)
income = rbind(income, income_nyc)

unemp2020_nyc = unemp2020 %>% 
  filter(fips %in% nyc) %>%  
  summarize(unemployed.rate =sum(unemployed) / (sum(unemployed) + sum(employed)))
unemp2020_nyc$fips = "nyc"
unemp2020_nyc$year = "2020"
unemp2020_nyc = unemp2020_nyc [,c(2,1,3)]
unemp2020 = unemp2020[,c(1,4,5)] %>% 
  filter(!fips %in% nyc)
unemp2020 = rbind(unemp2020, unemp2020_nyc)

unemp2021_nyc = unemp2021 %>% 
  filter(fips %in% nyc) %>%  
  mutate(employed = str_replace_all(employed, ",", "")) %>% 
  mutate(unemployed = str_replace_all(unemployed, ",", "")) %>% 
  mutate(employed = as.integer(employed), unemployed = as.integer(unemployed)) %>% 
  summarize(unemployed.rate =sum(unemployed) / (sum(unemployed) + sum(employed)))

unemp2021_nyc$fips = "nyc"
unemp2021_nyc$year = "2020"
unemp2021_nyc = unemp2021_nyc [,c(2,1,3)]
unemp2021 = unemp2021[,c(1,4,5)] %>% 
  filter(!fips %in% nyc)
unemp2021 = rbind(unemp2021, unemp2021_nyc)

voting_nyc = voting %>% 
  filter(fips %in% nyc) %>%  
  summarize(per_dem =sum(votes_dem) / sum(total_votes))
voting_nyc$fips = "nyc"
voting_nyc = voting_nyc [,c(2,1)]
voting= voting[,c(1,4)] %>% 
  filter(!fips %in% nyc)
voting = rbind(voting, voting_nyc)
```

```{r merge data}

employment = rbind(unemp2020, unemp2021)
info = unique(cases_deaths[,c(2,3,4)])

final_cases = cases_deaths %>% 
  mutate(peaks = case_when (date >= "2020-07-20" & date <= "2020-07-26" ~ "peak 1",
                            date >= "2021-01-08" & date <= "2021-01-14" ~ "peak 2",
                            date >= "2021-08-23" & date <= "2021-08-29" ~ "peak 3")) %>%
  filter(!is.na(peaks)) %>% 
  filter(fips!= "   NA") %>% 
  group_by(peaks, fips) %>% 
  summarise(cases = (max(cases) - min(cases))/7)

final = final_cases %>% 
  mutate(year = case_when(peaks == "peak 1" ~ "2020",
                          peaks == "peak 2" ~ "2021",
                          peaks == "peak 3" ~ "2021")) %>% 
  left_join(info, by = "fips") %>% 
  left_join(region, by = "state") %>% 
  left_join(geographic, by = "fips") %>% 
  left_join(voting, by = "fips") %>% 
  left_join(income, by = "fips") %>% 
  left_join(employment, by = c("year", "fips")) %>% 
  select(-year)

write.csv(final, "data/final.csv")

final %>% 
group_by(peaks) %>% 
summarise(cases = sum(cases))
```